\newpage
\TOCadd{Acknowledgements}

\begin{center}
ACKNOWLEDGEMENTS
\end{center}

\begin{flushright}
\textit{it was a long road...}
\\
Derek Church\\
\end{flushright}

Having been a graduate student for far too long, I have many thanks for many people.  To enumerate them all would produce a document of thesis-like length, so apologies in advance for anyone I miss from this list.

Firstly, my SUPERvisors: Dr. G. Tzanetakis saw a 3rd year undergraduate student (who was a foolishly shameful advocate of the C++ programming language) in his programming languages class and introduced him to the beautiful world of functional programming.  Dr. Y. Coady saw a geeky, shy 4th year student in her concurrency class and somehow convinced him to undertake graduate studies.  Both of them showed a seemingly limitless amount of patience and encouragement throughout my studies and I cannot thank them enough for this.

I also have to thank the far too many brilliant people at Business Objects (past and present) for both creating a wonderful programming environment, as well as for providing much inspiration along the way.  In particular, Luke Evans provided much of the early inspiration behind this thesis and I thank him for his input and insight.  Additionally, Kelly Booth and Davor Cubranic were an integral part of the project which formed much of the basis of this thesis, and I thank them for their input and support.  Additionally, funding for that project was provided by both Business Objects and the MITACS Accelerate program, so I am very much thankful for the financial support provided by them.

As well, one can not write any kind of acknowledgment without thanking family and friends.  I am fortunate to have many friends to thank, but Chris Ware in particular stands out as a colleague who was a ``sounding board'' on numerous occasions throughout my graduate studies.  In terms of family, I am truly grateful for the fact that both my Mother and my Father are still a part of my life, and have the opportunity to see me see this through (see Dad, I will finally finish school!).  And of course special thanks go to my wife Susan, who provided many hours, days, months, and years of support, encouragement, and kicks to the behind throughout my time at UVic.  Those who have been married while undertaking graduate studies will understand what I mean when I say that I never would have been able to complete my thesis without her and I am truly fortunate to have her in my life.

Last, but not least: \emph{42}.
